|

### 2021-05-21 04-Chp-CPRD-Analysis.Rmd
End of day: -3799

As a positive outcome control, we included ishemic heart disease. Statins have a known effect on this outcome on the basis of randomised controlled trials. __[CITATION NEEDED]__ We also used a negative outcome control of back pain to investigate the potential for residual confounding.

<!-- TODO need more information on how the controls were modelled - e.g. using the same covariates as in the main model or not? I.e. should those with baseline cardiovascular conditions be excluded from the IHD/MI analysis? -->

<!-- TODO Will also need some more here on why there were chosen -->

The inclusion of control outcomes are a useful technique to assess the strength of uncontrolled confounding in an analysis.[@lipsitch2010] Despite observational analyses suggesting a link between statins and muscle pain, systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the effect of statin use on muscle pain[@herrett2021] have found no evidence of an effect. As such, muscular backpain was used as a negative control outcome in this analysis.

<!----------------------------------------------------------------------------->
(ref:negativeOutcome-cap) Causal diagram showing relationship between exposure [X], outcome [Y], confounders (measured [C] and unmeasured [U]) and an ideal negative outcome [N]:

@@

<!-- TODO Need a lot more here - see other thesis for example of how covariates are described -->

The analysis was adjusted for a range of baseline covariates including sex,grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010), Charlson co-morbidity index, Index of Multiple Deprivation (IMD), consultation rate, alcohol (current, former,  never), smoking (current, former,  never), BMI, baseline total cholesterol, and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes.

All covariates were determined at index and definitions for each can be found in the archived data repository reflecting this study <!-- TODO link to Zenodo here -->

We ran two models, one unadjusted except for age, which was captured by using the participant's age as the time axis for the model,[@lamarca1998; @gail2009; @pencina2007] and a second adjusted for the additional covariates described above. Additional analyses stratified by outcome and drug class were also performed.

@@

<!-- TODO Need to complete paragraph below -->

@@

Missing data was only considered an issue for variables where a numerical test result was expected, (e.g. BMI) or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other conditions (e.g. statin use or dementia) is used to define a negative value of binary variables indicating the presence or absence of the condition, and so had no missing data at the point of imputation.[@wells2013strategies]

@@

As part of a Cox proportional hazard model, there is the option to use either absolute time or participants age as the time scale of interest. A model using age as the time axis inherently accounts, or adjusts, for participants age as a potential confounder of the exposure-outcome relationship.

In order to compare and contrast between the two approaches, we ran two simple model, one using participants age as the time-scale and a second using absolute time-since-entry and adjusting for participant age.

@@

@@

<!--------------------------------------------------------------------------->
(ref:immortalTimeBias-cap) __shortcap:__

@@

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures/sys-rev/prisma_flow.png"))
```
<!--------------------------------------------------------------------------->  
### Sensitivity analyses

@@

In addition to the primary analysis

Statins are contra-indicated in pregnant patients, and so I preformed a sensitivity analysis in patients aged 55 and coded as female.

We also prod

@@

@@

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) __Attrition of CPRD participants as the eligibility criteria were applied:__

@@

The median age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementia; <!-- see Supplementary Figure 1 -->). The distribution of baseline characteristics across the drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

@@

<!-- Table 1 is missing type 2 diabetes -->

@@

<!-- TODO Describe impact of SSA table - is the fact that many people stopped statin treatment (~10%) an issue? Comment specifically in the context of a -->

@@

Full covariate information was available for `r missingtext`. Five key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`, and baseline total cholesterol was missing for `r choltext`.

@@

@@

@@

@@

@@

<!-- TODO Sort out the mess that is below -->

@@

<!-- By multiple imputation versus complete case analysis -->

<!-- By pregnancy -->

<!-- By age-at-entry -->

<!-- By statin type (lipo vs hydrophilic) -->

<!-- By time varying confounder -->

@@

<!-- To what extent should I cross-reference with the other literature here? -->

@@

Lipid-regulating agents had no effect on probable and possible Alzheimer's when compared with no treatment, but were associated with increased risk of an all-cause dementia, vascular dementia and other dementia diagnosis. The effect observed in each case was driven by the statin subgroup, which included a substantial majority of participants. For the other drug classes, no association was found with any outcome, with two exceptions being that ezetimibe was associated with increased risk of vascular and other dementia, while fibrates were associated with increase risk of all-cause dementia and probable Alzheimer's disease.

@@

@@

<!----------------------------------------------------------------------------->
(ref:indicationBias-cap) __Confounding by indication directed acyclic graph:__

(ref:indicationBias-scap) Confounding by indication directed acyclic graph

@@

ggplot2::ggsave(filename = file.path("figures","cprd-analysis","indicationBias.png"),
dagitty::dagitty("dag {
  Statin [exposure,pos=\"0.000,0.000\"]
  \"Vascular outcomes\" [outcome,pos=\"2.000,0.000\"]
  \"Uncontrolled variable\" [pos=\"1.000,1.000\"]
  \"Uncontrolled variable\" -> Statin
  \"Uncontrolled variable\" -> \"Vascular outcomes\"
}") %>% ggdag_classic(size = 6) +
  xlim(-0.5,3) +
  geom_text(data = data.frame(x = 0.316392393234329,
                              y = 0.537705383514239,
                              label = "Prompts"),
            mapping = aes(x = x,
                          y = y,
                          label = label),
            angle = 0L,
            lineheight = 1L,
            hjust = 0.5,
            vjust = 0.5,
            colour = "red",
            family = "sans",
            fontface = "bold",
            inherit.aes = FALSE,
            show.legend = FALSE) +
  geom_text(data = data.frame(x = 1.65891420807966,
                              y = 0.537705383514239,
                              label = "Causes"),
            mapping = aes(x = x,
                          y = y,
                          label = label),
            angle = 0L,
            lineheight = 1L,
            hjust = 0.5,
            vjust = 0.5,
            colour = "red",
            family = "sans",
            fontface = "bold",
            inherit.aes = FALSE,
            show.legend = FALSE) +
  geom_curve(data = data.frame(x = 0.220612676359188,
                               y = -0.000788829298738025,
                               xend = 1.45,
                               yend = -0.00157828084913909),
             mapping = aes(x = x,
                           y = y,
                           xend = xend,
                           yend = yend),
             linetype = "dashed",
             angle = 0L,
             size = 0.6,
             colour = "black",
             curvature = 0,
             arrow = structure(list(angle = 0,
                                    length = structure(0,
                                                       class = "unit",
                                                       valid.unit = 2L,
                                                       unit = "inches"),
                                    ends = 2L,
                                    type = 2L),
                               class = "arrow"),
             inherit.aes = FALSE,
             show.legend = FALSE) +

@@

### Replicating other analytical strategies

@@

However, there was a large difference in the dementia "case mix", or proportion of different dementia diagnoses, between the two analyses, particularly in terms of the number of Alzheimer cases as a percentage of total dementia cases. The reason for this discrepancy between this analysis study and the previous paper is due to two things. While all of the codes used to define Alzheimer’s in the Smeeth paper are included in my “Probable Alzheimer’s” code-list (see Appendix \@ref(hold)), I also have several additional codes used to define this outcome (including, for example, “Eu00013: [X]Alzheimer's disease type 2”). Several of the codes in my “Possible Alzheimer’s” code list are included in the “Other dementia” code list used by Smeeth.

@@

<!-- TODO check that this is actually the case, that when you use the codelist from Smeeth, you get a different case-mix. Present it as a table. -->
<!-- Will need to present the results for this! -->
<!-- This, in addition to unexplained statistical issues such as adjusting for post-baseline covariates, means I am sceptical of the value of comparing our findings with Smeeth. -->
<!-- IN addition, this analysis used propensity s. IN addition, an analytical approach which appeared to adjust for key variables likely to be on the analytical pathway between   -->

@@

<!--------------------------------------------------------------------------->
(ref:smeethComparison-caption) Comparison of code lists used in this analysis and in Smeeth et al.

(ref:smeethComparison-scaption) Comparison of code lists used in this analysis and in Smeeth et al.

```{r smeethComparison-table, message=FALSE, results="asis", echo = FALSE}
```
<!--------------------------------------------------------------------------->

@@

 This chapter attempted to account for important sources of bias and provide a comparison with other available literature, as identified in the systematic review presented in Chapter \@ref(sys-rev-heading). However, there is a strong potential for unmeasured confounding, misclassification and reverse causation, which raises questions about our findings, in particular the unexpected increase in risk of vascular dementia associated with statin use. <!-- TODO Finish This is supported by our findings for the negative and positive controls used. A slight increas-->

Future research should aim to address these potential biases and, while it may be costly in terms of time and resources, a large scale, long-term randomized controlled trial would provide useful additional information on the effect of lipid-regulating agents on the risk of dementia and related outcomes.

<!-- TODO In my most recent discussion with supervisors, I discussed that a long term RCT isn't encouraged by the results -->

<!-- TODO Need better linkage with next chapter and triangulation -->

>>>>

In addition to our primary outcomes of interest, we created two additional control outcomes. The inclusion of control outcomes in observational analyses are a useful technique to assess the strength of uncontrolled confounding.[@lipsitch2010] Negative outcomes are those without a likely causal pathway between the exposure and outcome (see Figure \@ref()). Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, our confidence in the result for the outcome of interest is increased.

Muscular backpain was used as a negative control outcome in this analysis. Despite observational analyses suggesting a link between statins and muscle pain (as opposed to more serious complications such as myopathy), systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the effect of statin use on muscle pain[@herrett2021] have found no evidence of an effect.

Additionally, incident ischemic heart disease was included as a positive control outcome, given the well-established protective effect of lipid-lowering treatment on the risk of this condition.[@collins2016a]

<!-- QUESTION - should those with baseline cardiovascular conditions be excluded from the IHD/MI analysis? -->

<!----------------------------------------------------------------------------->
(ref:negativeOutcome-cap) Causal diagram showing relationship between exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of an arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders (assuming $C$ has been adjusted for).

@@

The analysis was adjusted for a range of baseline covariates. Demographic covariates included age and gender. Age was calculated at date of entry into the cohort. Socioeconomic status was proxied using the Index of Multiple Deprivation, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

Body mass index (calculated as $weight/height^2$), baseline total cholesterol and baseline LDL cholesterol measures were obtained, using the last recorded value prior to the index date. A variable indicating grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010) was included to allow for changes in prescribing trends across the lifetime of the cohort. To assess healthcare utilisation, I calculated the average annual number of consultations between the beginning of a patients data and their entry into the cohort

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new]

Codelists for all covariates can be found in the archived data repository accompanying this analysis (see Section \@ref(cprd-data-avail)).

@@

<!-- IDEA Add more detail here on the missing data approach -->

@@

Missing data was only considered an issue for variables where a numerical test result was expected, (e.g. BMI) or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition, rather than being considered missing.[@wells2013strategies]

@@

@@

<!-- TODO Need to describe accounting for clustering here: id it somethings I did? See other thesis for clustering by practice -->

As part of a Cox proportional hazard model, there is the option to use either absolute time in cohort or participants age as the time scale of interest.[@lamarca1998; @gail2009; @pencina2007] A model using age as the time axis inherently accounts, or adjusts, for participants age as a potential confounded of the exposure-outcome relationship. The main analyses presented all used age as the time axis.

In order to explore whether the choice of time axis affects the results of the analysis,I ran two simple models, one using participants age as the time-scale and a second using absolute time-since-entry and adjusting for participant age.

@@

<!--------------------------------------------------------------------------->
(ref:immortalTimeBias-cap) Diagram illustrating immortal time bias

@@

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures/sys-rev/prisma_flow.png"))
```
<!--------------------------------------------------------------------------->

To observe the effect of adjusting for additional covariates, I ran a model unadjusted except for age (captured via the time axis in the Cox model) and gender, and compared the results with the full adjusted model.

<!-- TODO Re-run this analysis -->

Sensitivity cohorts were also created, stratifying first by year of entry into the cohort and second by remiving participants who may have been pregnant (as statins are contraindicated in pregnancy).

Additional analyses stratified by outcome and drug class were also performed.

@@

@@

Exploring the results when using a different codelist

@@

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) Attrition of CPRD participants as the eligibility criteria were applied

@@

The median age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementia). The distribution of baseline characteristics across the drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

@@

<!-- FIX Table 1 is missing type 2 diabetes -->

@@

<!-- QUESTION Describe impact of SSA table - is the fact that many people stopped statin treatment (~10%) an issue? Comment specifically in the context of a -->

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`, and baseline LDL cholesterol was missing for `r ldltext`

@@

```{r p1Figure, include = FALSE}
```

@@

<!-- TODO Cross-check with the decision flowchart to ensure that  -->

<!-- TODO Check if reason for discrepnacy in terms of total numbers is that patients were excluded for having a first drug prior to on the index date -->

@@

__Choice of time axis__

__Complete case versus imputed data__

__Unadjusted versus full adjusted models__

__Sensitivity cohorts__

@@

__By statin type__

As detailed in the introduction, the properties of statins may be important in their effect, based on the ability of lip

__Control outcomes__

Using the fully adjusted model, the for the backpain negative control outcome was `r hold()` and the ischemic heart disease was `r hold()`

@@

<!-- TODO Sort out the mess that is **below** -->

@@

@@

<!-- QUESTION To what extent should I cross-reference with the other literature here? -->

@@

Lipid-regulating agents had a small effect on probable and possible Alzheimer's when compared with no treatment, but were associated with increased risk of an all-cause dementia, vascular dementia and other dementia diagnosis. The effect observed in each case was driven by the statin subgroup, which included a substantial majority of participants. For the other drug classes, no association was found with any outcome, with two exceptions being that ezetimibe was associated with increased risk of vascular and other dementia, while fibrates were associated with increase risk of all-cause dementia and probable Alzheimer's disease.

@@

<!-- TODO Discuss impact of consulation rate and effect on ascertainment bias - only adjusting for prior ascertainment rate, which could change once prescribed statins. However, would see increase effect across all sub-types, and ascertainment bias less likely for important high-impact events such as IHD -->

@@

<!----------------------------------------------------------------------------->
(ref:indicationBias-cap) Causal diagram (directed acyclic graph) illustrating confounding by indication

(ref:indicationBias-scap) Confounding by indication causal diagram

@@

ggplot2::ggsave(filename = file.path("figures","cprd-analysis","indicationBias.png"),
dagitty::dagitty("dag {
  Statin [exposure,pos=\"0.000,0.000\"]
  \"Vascular outcomes\" [outcome,pos=\"2.000,0.000\"]
  \"Uncontrolled variable\" [pos=\"1.000,1.000\"]
  \"Uncontrolled variable\" -> Statin
  \"Uncontrolled variable\" -> \"Vascular outcomes\"
}") %>% ggdag_classic(size = 6) +
  xlim(-0.5,3) +
  geom_text(data = data.frame(x = 0.316392393234329,
                              y = 0.537705383514239,
                              label = "Prompts"),
            mapping = aes(x = x,
                          y = y,
                          label = label),
            angle = 0L,
            lineheight = 1L,
            hjust = 0.5,
            vjust = 0.5,
            colour = "black",
            family = "sans",
            fontface = "bold",
            inherit.aes = FALSE,
            show.legend = FALSE) +
  geom_text(data = data.frame(x = 1.65891420807966,
                              y = 0.537705383514239,
                              label = "Causes"),
            mapping = aes(x = x,
                          y = y,
                          label = label),
            angle = 0L,
            lineheight = 1L,
            hjust = 0.5,
            vjust = 0.5,
            colour = "black",
            family = "sans",
            fontface = "bold",
            inherit.aes = FALSE,
            show.legend = FALSE) +
  geom_curve(data = data.frame(x = 0.220612676359188,
                               y = -0.000788829298738025,
                               xend = 1.45,
                               yend = -0.00157828084913909),
             mapping = aes(x = x,
                           y = y,
                           xend = xend,
                           yend = yend),
             linetype = "dashed",
             angle = 0L,
             size = 0.6,
             colour = "black",
             curvature = 0,
             arrow = structure(list(angle = 0,
                                    length = structure(0,
                                                       class = "unit",
                                                       valid.unit = 2L,
                                                       unit = "inches"),
                                    ends = 2L,
                                    type = 2L),
                               class = "arrow"),
             inherit.aes = FALSE,
             show.legend = FALSE) +

@@

### Use of alternative code lists

WHen using the Smeeth et al codelists were `r hold()` for Alzheimer's disease and `r hold()` for

However, direct comparison between the results is difficul

@@

However, there was a large difference in the dementia "case mix", or proportion of different dementia diagnoses, between the two analyses, particularly in terms of the number of Alzheimer cases as a percentage of total dementia cases. The reason for this discrepancy between this analysis study and the previous paper is due to two things. While all of the codes used to define Alzheimer’s in the Smeeth paper are included in my “Probable Alzheimer’s” code-list, I included several additional codes used to define this outcome (including, for example, “Eu00013: [X]Alzheimer's disease type 2”). Several of the codes in my “Possible Alzheimer’s” code list are included in the “Other dementia” code list used by Smeeth.

@@

<!-- TODO check that this is actually the case, that when you use the codelist from Smeeth, you get a different case-mix. Present it as a table. -->

<!-- Will need to present the results for this! -->

<!-- This, in addition to unexplained statistical issues such as adjusting for post-baseline covariates, means I am sceptical of the value of comparing our findings with Smeeth. -->

<!-- IN addition, this analysis used propensity s. IN addition, an analytical approach which appeared to adjust for key variables likely to be on the analytical pathway between   -->

@@

<!----------------------------------------------------------------------------->
(ref:smeethComparison-cap) Comparison of code used in this analysis with those used in the Smeeth et al. 2010 paper

(ref:smeethComparison-scap) Comparison of code used in this analysis with those used in the Smeeth et al. 2010 paper

```{r sankeydiagram, include=FALSE}
```

```{r smeethComparison, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:smeethComparison-cap)', out.width='100%', fig.scap='(ref:smeethComparison-scap)'}
knitr::include_graphics(file.path("figures/cprd-analysis/sankey_diagram.png"))
```
<!----------------------------------------------------------------------------->

@@

This chapter attempted to account for important sources of bias and provide a comparison with other available literature, as identified in the systematic review presented in Chapter \@ref(sys-rev-heading). However, there is a strong potential for unmeasured confounding by indication and differential misclassification of the outcome on the basis of exposure, which raises questions about our findings, in particular the unexpected increase in risk of vascular dementia associated with statin use. This is supported by our findings for the negative and positive control outcomes used. Future research using large scale electronic health records should aim to address these potential biases, potentially using a analytical design that more closely emulates a trial.[@danaei2013b]

The unexpected increase in vascular dementia risk with statin use is particularly interesting given the absence of vascular dementia in the published literature, as highlighted by the previous chapter (see Section \@ref(sys-rev-pub-bias)) . It is possible that previous research identified a similar effect and via a publication bias mechanism, these results did not make it into the evidence base.

Regardless, this analysis has provided an additional source of evidence for the triangulation exercise presented in Chapter \@ref(discussion-heading).

In the following Chapter, the dataset described here is incorporated along with several other datasets
